Working from home and digital divides: resilience during the pandemic

### 2021-03-26 paper/v2_taylor_francis/broadband.speed.covid_anonymised.tex
v.2
second submission

\author{\name{Hannah Budnitz$^{a}$, Emmanouil Tranos$^{b, c}$}
\affil{$^{a}$Transport Studies Unit, University of Oxford, UK; $^{b}$School of Geographic Sciences, University of Bristol, UK; $^{c}$The Alan Turing Institute, UK}
}

\thanks{CONTACT Hannah Budnitz. Email: , Emmanouil Tranos. Email: \href{mailto:e.tranos@bristol.ac.uk}{\nolinkurl{e.tranos@bristol.ac.uk}}}

@@

During the pandemic, working from home using digital technologies,
whether partially or exclusively, was transformed from a niche means of
accessing work, albeit one that had been on a slow, upward trend, to a
widespread way of life in many countries. The ability to work from home
or telecommute meant millions retained their jobs and, to a varying
extent, maintained productivity during periods of strict lockdowns
around the world. However, this ability has not been evenly distributed
socially or spatially, creating new intersections of economic and
digital divisions. On one side are those who can work from home,
supported by digital technologies, and have thus been able to enjoy both
economic resilience and greater personal safety. On the other side,
previously employed individuals have been forced to accept furlough or
redundancy packages unless they are part of the cadre of essential
workers, who are potentially at high risk of infection. Whilst the basis
for this pandemic-generated divide has been viewed as mainly
occupational, here we consider whether it is also technological and,
consequently, geographical.

Using the UK as a case study, this paper aims to understand how the
quality and reliability of internet service, as reflected in
\emph{experienced} internet speeds during the spring 2020 lockdown, may
reinforce or redress the spatial and social dimensions of digital
divisions. We employ volunteered geographic data on individual broadband
speed tests and state-of-the-art time-series clustering methods to
create clusters of UK local authorities with similar temporal signatures
of experienced internet speeds. We then associate these clusters of
local authorities with their socioeconomic and geographic
characteristics to explore how they overlap with or diverge from the
existing economic and digital geography of the UK. Our analysis enables
us to better understand how the spatial and social distribution of
education, occupation and online accessibility intersect to enable or
hinder the practice of telecommuting at a time of extreme demand. We
also consider what lessons can be learned from this time for a future
where telecommuting is likely to remain a more common means of accessing
work, at least in comparison to the pre-Covid era.

The capability to work from home has previously been studied from the
perspective of whether work tasks in a given occupation both can be and
are allowed to be performed using digital technologies independently of
location or co-location with colleagues, including supervisors
\citep{allen2015effective, singh2013modeling}. However, successful
telecommuting also requires that the quality and reliability of digital
services, particularly home internet connection speeds, enable the
completion of work tasks with a minimum of delay or interruption. Prior
to the pandemic, the performance of broadband services with respect to
telecommuters was never tested at scale, as working from home and
connecting to colleagues and workplace resources via the internet was
the purview of a small minority of workers. Instead, leisure use in the
evening, when video streaming services are at their peak, has been used
to benchmark broadband performance and service delivery by different
Internet Service Providers (ISPs) \citep{ofcom2017}.

Yet the shift towards telecommuting during various stages of lockdown
around the world has been drastic and there are speculations that
post-Covid, the tendency to work from home will be much higher, raising
questions around whether internet services can accommodate the increased
demand. For example, \(47\)\% of people in employment in the UK worked
solely from home in April \(2020\), whilst the same figure only reached
\(5\)\% the year before \citep{ons2020, ons2020lm2019}. A back of the
envelope calculation suggests that up to 40\% of the labour force could
work from home on an ongoing basis \citep{batty2020editorial}. Similar
figures have been reported for other countries
\citep{felstead2020homeworking}. Approximately \(37\)\% of the European
workforce worked from home in April \(2020\) with countries like Finland
reaching \(60\)\% \citep{eurofound2020}. In the US, almost half of the
working population worked from home during the same period because of
the pandemic \citep{brynjolfsson2020covid}, and a recent estimate
indicated that \(37\)\% of all jobs in the US can be permanently
performed entirely from home \citep{NBERw26948}. None of these changes
could have happened in the absence of reliable information and
communication technology (ICT) infrastructure -- both in terms of
software and hardware. But while software innovations are easily
diffused across space and society\footnote{See for example the huge
  success of videoconferencing apps such as Zoom \citep{marks2020zoom}.},
the same does not apply for ICT hardware infrastructure such as internet
broadband connectivity.

As the quality of internet infrastructure and services, as well as the
concentration of different occupations are spatially dependent and
clustered in space, our approach offers a framework for understanding
the impact of and interactions between digital divisions geographically
and socioeconomically. By asking how resilient broadband speeds, and
particularly upload speeds, are as experienced in different parts of the
UK during a time of extreme demand, we interrogate which places benefit
from the greater economic resilience digital technologies can offer. The
structure of this paper is as follows. First we review the literature on
telecommuting and digital divides to better understand their structural
and spatial development pre-pandemic, and thus their importance to the
economic resilience of different places. We then describe our data and
methodology. Our results section first offers a classification of how
internet services vary across clusters of UK local authorities and then
assesses whether these clusters replicate or repudiate other
socio-economic and geographic patterns of economic resilience. We
conclude with a discussion of the insights we have gained from our new
perspective on digital divisions.

@@

Our understanding of telecommuting as a product of enabled occupations
can be described as a manifestation of the third level digital divide,
as those who are able to use digital technologies to work from home
benefit from a high rate of return on their use of the internet in terms
of autonomy, flexibility, and time saved from commuting
\citep{peters2004employees, siha2006telecommuting, singh2013modeling}.
These returns have been even greater during the Covid-19 crisis, when
those with the capability to telecommute also have the ability to
maintain their employment whilst protecting their health. However, the
success of these arrangements has been dependent upon the first level
digital divide, which is associated with access to and quality of
internet connectivity. \citet{SALEMINK2017360} provides a systematic
review of the pre-pandemic, first level digital divide in infrastructure
quality between urban and rural areas in various advanced economies.
Rural areas, predictably, fare worse, yet as \citet{blank2018local}
highlight, variation in individual internet uptake and use is a product
of more complex spatial and demographic characteristics than simple
rurality or urbanisation. Yet whether this variation in infrastructure
quality affects the spatial footprint of telecommuting has not
previously been investigated, in part because telecommuting has not
previously been a cause of concentrated demand and pressure on internet
services.

There are indications that those who purchase high speed connections
consume more data of all sorts and use their connections for a greater
variety of purposes \citep{hauge2011consumer}. There is also a
correlation between access to internet services and a reduction in
household transport spend \citep{bris2017ict}. Independently of whether
the additional internet use and reduced travel is because of increased
telecommuting, these studies suggest that better internet services
enable households to make savings and efficiencies, an example of the
first level digital divide reinforcing the third level.

Multi-layered digital divides may also intersect with material divides
and the economic geography of the UK. The regional economic resilience
literature underlines the differentiated capacity of cities and regions
to escape or recover from economic crises
\citep{martin2012regional, kitsos2018economic}. As different places have
different industrial and occupational profiles, these affect their
aggregated potential capacity for telecommuting. Such profiles are
associated with longstanding inequalities in the UK and their spatial
representation as a North-South divide \citep{martin_north_south}.
Various studies have illustrated severe inequalities between the north
and the south regions of England in terms of skills and human capital,
unemployment, productivity and prosperity
\citep{lee2014grim, mccann2020perceptions, dorling2018peak}. Some
scholars have even argued that the UK suffers from some of the highest
levels of interregional inequalities in the global north
\citep{gal2018reducing, mccann2016uk}. All three levels of digital
divides are associated with or shaped by the geography of the UK. Yet
this is the first time that the intersection of digital and material
divides is relevant to understanding the economic resilience of places
and large swathes of the population, as digital technologies became an
essential tool of productivity during the Covid-19 pandemic.

The extreme demand during the pandemic thus provides a new opportunity
to understand how internet infrastructure quality, and reliability
affects telecommuting, particularly in light of the high volumes of
bandwidth-intensive video conferencing required in order to avoid the
face-to-face contact that could increase the spread of infection. We
seek to answer how internet service resilience could contributes to or
reduce economic resilience when the latter is dependent upon the
capability to work from home. We also aim to improve our understanding
of the impact of first level digital divisions on telecommuting, and
whether this results in more fundamental third level digital divisions
than has previously been perceived.

@@

To assess the internet quality and reliability across local authorities
in the UK, we utilise unique data comprising individual internet speed
tests from Speedchecker Ltd\footnote{\url{https://www.broadbandspeedchecker.co.uk/}}.
This is a private company that allows internet users to check their own
broadband upload and download speeds, and stores every speed-check with
timestamp and geolocation information. These data have been used before
to assess digital divides \citep{riddlesden2014broadband} and the impact
of local loop unbundling regulatory processes
\citep{nardotto2015unbundling}, and we followed the former's approach to
remove outliers. By using this volunteered geographic data, we are able
to assess the internet speed \emph{experienced} by users, which may
differ from the maximum speeds \emph{advertised} by ISPs. We are
particularly interested in upload speeds and the frequency of speed
tests over the period from March to May \(2020\), as government
statements indicate this is when UK workers were first told to work from
home if at all possible \citep{GovUK2020}. Average upload speeds are
slower than average download speeds, at \(9.3\)Mb/s mean upload speed
for the whole sample, compared to \(29.6\)Mb/s for download speeds, but
they are also less associated with internet-based, high-demand, leisure
activities such as video streaming. Therefore, upload speeds are more
relevant to work-related activities such as uploading documents or
two-way audio, video, and text-based communication systems.

The frequency of speed tests was important in identifying the temporal
profile which would give us most insight into experienced internet
service and resilience over units of time. Whilst there is an overall
trend of increased testing from March to April and then a slight
reduction from April to May, this trend masks substantial variation by
not only the day of the week, but also time of day, as can be seen in
Figure \ref{test2020}. Thus, a daily aggregation of upload speeds would
mask the variation in experienced service over the course of each
weekday. Furthermore, the importance of this variation is highlighted by
a comparison with the same period in \(2019\), as in Figure
\ref{test2019}, when the volume of testing and thus of experienced
internet service quality peaked in the evening, presumably in response
to demand for leisure activities and download speeds. In contrast, the
majority of the increase in testing in \(2020\) is during the working
day, creating a new morning peak in Figure \ref{test2020}. Therefore, we
include a measure of hourly variation in our temporal profiles to
reflect the change in users' perception of the workday reliability of
internet services.

However, there were insufficient observations -- only \(631\) speed
tests per LAD on average -- for each for each working hour of each
working day in each Local Authority District (LAD) to profile speeds at
that level of detail. Spatial aggregation was also necessary because we
could not follow individuals or households and connect data points.
Therefore, we aggregate the \(241,088\) individual, geolocated and
time-stamped speed-checks during the \(13\) weeks of March to May
inclusive for weekdays in \(2020\) by each hour of the day and day of
the week. As our research aims to identify the geography of internet
service resilience for work purposes, bank holidays and the hours
between midnight and \(6:00\) were excluded, as well as weekend days.
The composite week time-series thus comprise \(18\) hours multiplied by
\(5\) weekdays or \(90\) time points per series. The time-series were
calculated for each of the \(382\) LADs in the UK, standardised, and
then a \emph{k}-means partitioning around medoids clustering algorithm
was applied using DTW. We initially run the algorithm for
\(k \in \mathbb{N} \bigcap [5,15]\) and used cluster validity indices
(CVIs) to pick the optimal solution of \(k = 13\). Following
\citet{sardatime} the majority vote for the following CVIs was used:
Silhouette (max), Score function (max), Calinski-Harabasz (max),
Davies-Bouldin (min), Modified Davies-Bouldin (DB*, min), Dunn (max),
COP (min).

@@

The temporal profiles of the local authority clusters have been
summarised in Figures \ref{UpClusterL} and \ref{UpClusterS} and Table
\ref{up.cluster.descr}. The graphs show a composite profile of mean
upload speeds per hour per day for each cluster, with the largest five
clusters, in terms of the LAD membership and population, in Figure
\ref{UpClusterL}, and the next six in Figure \ref{UpClusterS}. These
figures and table provide a comprehensive overview of the quality and
reliability of experienced broadband in different parts of the UK.

\begin{figure}
\includegraphics[width=0.95\linewidth]{figures/upClusterL} \caption{\label{UpClusterL}Temporal profilies for upload speed large clusters}\label{fig:unnamed-chunk-2}
\end{figure}

\begin{figure}
\includegraphics[width=0.95\linewidth]{figures/upClusterS} \caption{\label{UpClusterS}Temporal profilies for upload speed small clusters}\label{fig:unnamed-chunk-3}
\end{figure}

\begin{table}[!htbp] \centering 
  \caption{Upload speed cluster characteristics\label{up.cluster.descr}} 
  \label{} 
\footnotesize 
\begin{tabular}{@{\extracolsep{0pt}} ccccccc} 
\\[-1.8ex]\hline 
\hline \\[-1.8ex] 
Cluster & N. of LADs & LAD population & mean speed & SD speed & mean AM speed & mean PM speed \\ 
\hline \\[-1.8ex] 
1 & 5 & 343100 & 8557 & 6139 & 7747 & 9563 \\ 
2 & 2 & 265600 & 10922 & 6687 & 9674 & 10645 \\ 
3 & 4 & 474700 & 10201 & 5658 & 9470 & 11236 \\ 
4 & 1 & 91100 & 9689 & 6122 & 7816 & 9689 \\ 
5 & 1 & 79800 & 10127 & 6024 & 9030 & 11101 \\ 
6 & 155 & 29535700 & 9397 & 5839 & 9161 & 9580 \\ 
7 & 4 & 559800 & 10119 & 6102 & 9813 & 11070 \\ 
8 & 5 & 436300 & 9429 & 6254 & 8682 & 10434 \\ 
9 & 32 & 6355500 & 10878 & 5957 & 10832 & 11071 \\ 
10 & 4 & 699600 & 10795 & 6005 & 9258 & 10697 \\ 
11 & 33 & 5771400 & 10845 & 5936 & 10781 & 10988 \\ 
12 & 10 & 1544900 & 9551 & 6166 & 9254 & 9048 \\ 
13 & 126 & 20277700 & 8392 & 5849 & 8299 & 8522 \\ 
\hline \\[-1.8ex] 
\multicolumn{7}{l}{Note: All speed measures are upload speeds} \\ 
\end{tabular} 
\end{table}

The second largest cluster, comprising \(126\) local authorities and
over \(20\) million people, is cluster \(13\), which has the slowest
aggregate mean upload speed, and the second highest ratio of the
standard deviation to the mean. This suggests that those living in local
authorities in this cluster experienced some of the lowest quality
broadband services in terms of upload speeds in the UK. However, Figure
\ref{UpClusterL} shows that the high standard deviation, which is one
indication of unreliable internet, did not disproportionately affect the
morning peak from \(9:00\)-\(10:59\), when upload speeds were, on
average, only \(2.6\)\% slower than in the evening peak period between
\(19:00\) and \(20:59\), when entertainment purposes are likely to be
using the most bandwidth. In comparison, the five LADs in cluster \(1\),
home to \(343\) thousand people, not only experience the second slowest
mean upload speeds and the highest ratio of standard deviation to the
mean, but are also much more affected during the morning peak.

Meanwhile, those living in the largest cluster -- \(6\), with \(155\)
LADs and \(29.5\) million people -- experienced aggregate mean upload
speeds of about \(1\)Mb/s faster than those in cluster \(13\), but still
lower than the other three large clusters and most of the smaller
clusters, suggesting a middling quality of service. The temporal profile
for cluster \(6\) in Figure \ref{UpClusterL} shows that upload speeds
are highest at \(6:00\) on a Monday morning and between \(23:00\) and
midnight on Wednesday and Thursday, but tend to be lower during the
working day. Furthermore, experienced mean upload speeds in the morning
peak are \(4.4\)\% lower than in the evening peak -- a greater, more
noticeable change than any of the other large clusters experience,
suggesting poorer reliability during the working day. This difference is
less, however than any of the clusters included in Figure
\ref{UpClusterS}.

Clusters \(8\) and \(12\) also have mean upload speeds under \(10\)Mb/s,
but higher than clusters \(1\), \(13\), and \(6\). However, mean upload
speeds are much lower between \(9:00-10:59\) than between
\(19:00-20:59\) in cluster \(8\), but slightly faster in the morning
than in the evening in cluster \(12\). Indeed, cluster \(12\) is the
only cluster to experience higher speeds in the morning peak compared to
the evening, and thus the only cluster where the temporal profile of
internet use is closer to what might have been expected pre-pandemic.
Among the other clusters, however, the reliability of internet services
during the working day still varies considerably. Interpreting this
variation from the large spikes and dips shown on Figures
\ref{UpClusterL} and \ref{UpClusterS} is difficult, but the statistics
in Table \ref{up.cluster.descr} show that clusters \(9\) and \(11\) have
the most reliable internet services. The ratio of standard deviation to
mean in both these clusters is lowest, and speeds are only about 2\%
slower in the morning than the evening. Mean speeds are also higher than
in any other cluster, excluding cluster \(2\), where measures of
reliability suggest poorer performance.

>>>>

\author{\name{$^{a}$}
\affil{$^{a}$}
}

\thanks{CONTACT . Email: }

@@

During the pandemic, working from home using digital technologies,
whether partially or exclusively, was transformed from a niche means of
accessing work, albeit one that had been on a slow, upward trend, to a
widespread way of life in many countries. The ability to work from home
or telecommute meant millions retained their jobs and, to a varying
extent, maintained productivity during periods of strict lockdowns
around the world. However, this ability has not been evenly distributed
socially or spatially, creating new intersections of economic and
digital divisions. On one side are those who can work from home,
supported by digital technologies, and have thus been able to enjoy both
economic resilience and greater personal safety. On the other side,
previously employed individuals have been forced to accept furlough or
redundancy packages unless they are part of the cadre of essential
workers, who are potentially at high risk of infection. Whilst the basis
for this pandemic-generated divide has been viewed as mainly
occupational, here we consider whether it is also technological and
geographical.

The capability to work from home has previously been studied from the
perspective of whether work tasks in a given occupation both can be and
are allowed to be performed using digital technologies independently of
location or co-location with colleagues, including supervisors
\citep{allen2015effective, singh2013modeling}. However, successful
telecommuting also requires that the quality and reliability of digital
services, particularly home internet connection speeds, enable the
completion of work tasks with a minimum of delay or interruption. High
levels of demand are one of the main causes of reduced reliability and
slower connection speeds, as network bandwidth becomes congested. Yet
prior to the pandemic, the performance of broadband services with
respect to telecommuters was never tested at scale, as working from home
and connecting to colleagues and workplace resources via the internet
was the purview of a small minority of workers. Instead, leisure use in
the evening, when video streaming services are at their peak, has been
used to benchmark broadband performance and service delivery by
different Internet Service Providers (ISPs) \citep{ofcom2017}.

The shift towards telecommuting during various stages of lockdown around
the world has been drastic and there are speculations that post-Covid,
the tendency to work from home will be much higher, raising questions
around whether internet services can accommodate the increased demand.
For example, \(47\)\% of people in employment in the UK worked solely
from home in April \(2020\), whilst the same figure only reached \(5\)\%
the year before \citep{ons2020, ons2020lm2019}. Similar figures have
been reported for other countries \citep{felstead2020homeworking}.
Approximately \(37\)\% of the European workforce worked from home in
April \(2020\) with countries like Finland reaching \(60\)\%
\citep{eurofound2020}. In the US, almost half of the working population
worked from home during the same period because of the pandemic
\citep{brynjolfsson2020covid}, and a recent estimate indicated that
\(37\)\% of all jobs in the US can be permanently performed entirely
from home \citep{NBERw26948}. None of these changes could have happened
in the absence of reliable information and communication technology
(ICT) infrastructure -- both in terms of software and hardware. But
while software innovations are easily diffused across space and
society\footnote{See for example the huge success of videoconferencing
  apps such as Zoom \citep{marks2020zoom}.}, the same does not apply for
ICT hardware infrastructure such as internet broadband connectivity.

Using the UK as a case study, this paper aims to understand how the
quality and reliability of internet service, as reflected in
\emph{experienced} internet speeds during the spring 2020 lockdown, may
reinforce or redress the spatial and social dimensions of digital
divisions. We employ volunteered geographic data on individual broadband
speed tests and state-of-the-art time-series clustering methods to
create clusters of UK local authorities with similar temporal signatures
of experienced internet speeds. We then associate these clusters of
local authorities with their socioeconomic and geographic
characteristics to explore how they overlap with or diverge from the
existing economic and digital geography of the UK. As the quality of
internet infrastructure and services, as well as the concentration of
different occupations are spatially dependent and clustered in space,
our approach offers a framework for understanding the impact of and
interactions between digital divisions geographically and
socioeconomically. Our analysis enables us to better understand how the
spatial and social distribution of occupation and online accessibility
intersect to enable or hinder the practice of telecommuting at a time of
extreme demand. We also consider what lessons can be learned from this
time for a future where telecommuting is likely to remain a more common
means of accessing work, at least in comparison to the pre-Covid era.

By asking how resilient broadband speeds, and particularly upload
speeds, are as experienced in different parts of the UK during a time of
extreme demand, we interrogate which places benefit from the greater
economic resilience digital technologies can offer. The structure of
this paper is as follows. First we review the literature on
telecommuting and digital divides to better understand their structural
and spatial development pre-pandemic, and thus their importance to the
economic resilience of different places. We then describe our data and
methodology. Our results section first offers a classification of how
internet services vary across clusters of UK local authorities and then
assesses whether these clusters replicate or repudiate other
socio-economic and geographic patterns of economic resilience. We
conclude with a discussion of the insights we have gained from our new
perspective on digital divisions.

@@

Our understanding of telecommuting as a product of enabled occupations
can be described as a manifestation of the third level digital divide,
as those who are able to use digital technologies to work from home
benefit from a high rate of return on their use of the internet in terms
of autonomy, flexibility, and time saved from commuting
\citep{peters2004employees, siha2006telecommuting, singh2013modeling}.
In some aspects, these returns may have increased during the Covid-19
crisis, when those with the capability to telecommute also have the
ability to maintain their employment whilst protecting their health.
However, the success of these arrangements has been dependent upon the
first level digital divide, which is associated with access to and
quality of internet connectivity. \citet{SALEMINK2017360} provides a
systematic review of the pre-pandemic, first level digital divide in
infrastructure quality between urban and rural areas in various advanced
economies. Rural areas, predictably, fare worse. This finding broadly
applies in the UK, where studies that also analysed broadband speed
checks concluded that average speeds are lower in rural areas, something
that has not been improved by policy measures to increase competition
\citep{riddlesden2014broadband, nardotto2015unbundling}. In contrast,
\citet{riddlesden2014broadband} found that levels of deprivation did not
correlate with first level digital divides.

However, as \citet{blank2018local} highlight, variation in individual
internet availability and uptake is a product of more complex spatial
and demographic characteristics than simple rurality or urbanisation.
Dense urban areas were shown to suffer more from slowdown during peak
hours, although these services were more likely to be improved by
increased competition between providers, such as between new entrants
and Virgin Media cable connections
\citep{riddlesden2014broadband, nardotto2015unbundling}. The latter were
historically available to only \(45\)\% of premises in the UK
\citep{ofcom2016}, where the more lucrative and competitive market
originally attracted the cable TV provider. Whether the variation in
infrastructure quality and reliability affects the spatial footprint of
telecommuting has not previously been investigated, in part because
telecommuting has not previously been a cause of concentrated demand and
pressure on internet services.

Other intersections between digital divides have been subject to study.
There are indications that those who purchase high speed connections
consume more data of all sorts and use their connections for a greater
variety of purposes \citep{hauge2011consumer}. There is also a
correlation between access to internet services and a reduction in
household transport spend \citep{bris2017ict}. These studies suggest
that better internet services enable households to make savings and
efficiencies, an example of the first level digital divide reinforcing
the third level. Such digital intersections and efficiencies have also
changed transportation, connectivity and the structure of cities during
the pandemic. We now have enough data to confirm the drastic alteration
observed in the space-time geography of cities around the world in the
last year \citep[e.g.~][]{google2020, shibayama2021impact}. Yet,
although there is a broad agreement that these changes during the
pandemic have played a pivotal role in stopping the spread of the virus
\citep{jia2020population, yang2020taking, mu2020interplay}, the extent
to which the increased levels of working from home, the consequent
decrease in commuting flows, and the altered structure of cities will
remain post-pandemic is the subject of considerable debate in the
literature.

Meanwhile, multi-layered digital divides may also intersect with
material divides and the economic geography of the UK. The regional
economic resilience literature underlines the differentiated capacity of
cities and regions to escape or recover from economic crises
\citep{martin2012regional, kitsos2018economic}. As different places have
different industrial and occupational profiles, these affect their
aggregated potential capacity for telecommuting. Such profiles are
associated with longstanding inequalities in the UK and their spatial
representation as a North-South divide \citep{martin_north_south}.
Various studies have illustrated severe inequalities between the north
and the south regions of England in terms of skills and human capital,
unemployment, productivity and prosperity
\citep{lee2014grim, mccann2020perceptions, dorling2018peak}. Some
scholars have even argued that the UK suffers from some of the highest
levels of interregional inequalities in the global north
\citep{gal2018reducing, mccann2016uk}. All three levels of digital
divides are associated with or shaped by the geography of the UK. Yet
this is the first time that the intersection of digital and material
divides is relevant to understanding the economic resilience of places
and large swathes of the population, as digital technologies became an
essential tool of productivity during the Covid-19 pandemic.

The extreme demand during the pandemic thus provides a new opportunity
to understand how internet infrastructure quality, and reliability
affects telecommuting, particularly in light of the high volumes of
bandwidth-intensive video conferencing required in order to avoid the
face-to-face contact that could increase the spread of infection. We
seek to answer how internet service resilience could contribute to or
reduce economic resilience when the latter is dependent upon the
capability to work from home. We also aim to improve our understanding
of the impact of first level digital divisions on telecommuting, and
whether this results in more fundamental third level digital divisions
than has previously been perceived.

@@

To assess the internet quality and reliability across local authorities
in the UK, we utilise unique data comprising individual internet speed
tests from Speedchecker Ltd\footnote{\url{https://www.broadbandspeedchecker.co.uk/}}.
This is a private company that allows internet users to check their own
broadband upload and download speeds, and stores every speed-check with
timestamp and geolocation information. These data have been used before
to assess digital divides \citep{riddlesden2014broadband} and the impact
of local loop unbundling regulatory processes
\citep{nardotto2015unbundling}, and we followed the former's approach to
remove outliers. By using this volunteered geographic data, we are able
to assess the internet speed \emph{experienced} by users, which may
differ from the maximum speeds \emph{advertised} by ISPs. Indeed, those
who seek to test their broadband are most likely to do so because they
are experiencing slower speeds than expected, although any skew towards
slower speeds is balanced by the likelihood that those who test their
broadband are also more `tech-savvy' and / or have purchased higher
speed packages that are not delivering the promised level of service.

We are particularly interested in upload speeds and the frequency of
speed tests over the period from March to May \(2020\), as government
statements indicate this encompasses the period when UK workers were
told to work from home if at all possible \citep{GovUK2020}. Schools and
various retail, leisure and hospitality businesses were closed from late
March, and restrictions were gradually eased from late May. Average
upload speeds are slower than average download speeds, at \(9.3\)Mb/s
mean upload speed for the whole sample, compared to \(29.6\)Mb/s for
download speeds, but they are also less associated with internet-based,
high-demand, leisure activities such as video streaming. Therefore,
upload speeds are more relevant to work-related activities such as
uploading documents or two-way audio, video, and text-based
communication systems.

The frequency of speed tests was important in identifying the temporal
profile which would give us most insight into experienced internet
service and resilience over units of time. Whilst there is an overall
trend of increased testing from March to April and then a slight
reduction from April to May, this trend masks substantial variation by
not only the day of the week, but also time of day, as can be seen in
Figure \ref{test2020} \footnote{Figures \ref{test2020} and
  \ref{test2019} were created with the \texttt{R} package
  \texttt{openair} \citep{openair}.}. Thus, a daily aggregation of
upload speeds would mask the variation in experienced service over the
course of each weekday. Furthermore, the importance of this variation is
highlighted by a comparison with the same period in \(2019\), as in
Figure \ref{test2019}, when the volume of testing and thus of
experienced internet service quality peaked in the evening, presumably
in response to demand for leisure activities and download speeds. In
contrast, the majority of the increase in testing in \(2020\) is during
the working day, creating a new morning peak in Figure \ref{test2020}.
Therefore, we include a measure of hourly variation in our temporal
profiles to reflect the change in users' perception of the workday
reliability of internet services.

However, there were insufficient observations -- only \(631\) speed
tests per Local Authority District (LAD) on average -- for each for each
working hour of each working day in each LAD to profile speeds at that
level of detail. Spatial aggregation was also necessary because we could
not follow individuals or households and connect data points. Therefore,
we aggregate the \(241,088\) individual, geolocated and time-stamped
speed-checks during the \(13\) weeks of March to May inclusive for
weekdays in \(2020\) by each hour of the day and day of the week. As our
research aims to identify the geography of internet service resilience
for work purposes, bank holidays and the hours between midnight and
\(6:00\) were excluded, as well as weekend days. The composite week
time-series thus comprise \(18\) hours multiplied by \(5\) weekdays or
\(90\) time points per series. The time-series were calculated for each
of the \(382\) LADs in the UK, standardised, and then a \emph{k}-means
partitioning around medoids clustering algorithm was applied using DTW.
We initially run the algorithm for \(k \in \mathbb{N} \bigcap [5,15]\),
calculated the cluster validity indices (CVIs) and then run the
subsequent multinomial regression -- see the end of this section for
more details for the different \emph{k}. Following \citet{sardatime}, to
identify the optimal \emph{k} we used the majority vote for the
following CVIs: Silhouette (max), Score function (max),
Calinski-Harabasz (max), Davies-Bouldin (min), Modified Davies-Bouldin
(DB*, min), Dunn (max), COP (min). Nevertheless, we opted against using
the optimal \(k = 13\) solution as it was too large to allow for
communicable LAD clusters. Instead we opted for a smaller \(k = 9\),
which led to a rather similar spatial pattern and, importantly, to a
much higher R-squared in the subsequent explanatory regression (\(0.44\)
instead of \(0.34\)).

@@

The temporal profiles of the local authority clusters have been
summarised in Figure \ref{UpCluster} and Table \ref{up.cluster.descr},
to provide an overview of the quality and reliability of experienced
broadband in different parts of the UK. Figure \ref{UpCluster} shows a
composite profile of mean upload speeds per hour per day for each of the
largest six clusters, in terms of the LAD membership and population.

\begin{figure}
\includegraphics[width=0.95\linewidth]{figures/UpCluster} \caption{\label{UpCluster}Temporal profilies for upload speed clusters}\label{fig:unnamed-chunk-2}
\end{figure}

\begin{table}[!htbp] \centering 
  \caption{Upload speed cluster characteristics\label{up.cluster.descr}} 
  \label{} 
\footnotesize 
\begin{tabular}{@{\extracolsep{0pt}} ccccccc} 
\\[-1.8ex]\hline 
\hline \\[-1.8ex] 
Cluster & N. of LADs & LAD population & mean speed & SD speed & mean AM speed & mean PM speed \\ 
\hline \\[-1.8ex] 
1 & 9 & 903200 & 9564 & 6314 & 8798 & 10457 \\ 
2 & 2 & 162000 & 12085 & 6537 & 11882 & 10866 \\ 
3 & 12 & 1785800 & 11047 & 6079 & 10029 & 11634 \\ 
4 & 1 & 91100 & 9689 & 6122 & 7816 & 9689 \\ 
5 & 3 & 280000 & 10802 & 6116 & 11010 & 10084 \\ 
6 & 229 & 40552800 & 8761 & 5847 & 8555 & 8955 \\ 
7 & 5 & 682500 & 10326 & 6102 & 10045 & 11149 \\ 
8 & 6 & 510000 & 9769 & 6352 & 8989 & 10836 \\ 
9 & 115 & 21467800 & 10328 & 5915 & 10283 & 10333 \\ 
\hline \\[-1.8ex] 
\multicolumn{7}{l}{Note: All speed measures are upload speeds} \\ 
\end{tabular} 
\end{table}

The largest cluster, comprising \(229\) local authorities and over
\(40\) million people, is cluster \(6\), which has the slowest aggregate
mean upload speed of any cluster, and the highest ratio of the standard
deviation to the mean. This suggests that those living in local
authorities in this cluster experienced some of the lowest quality
broadband services in terms of upload speeds and reliability in the UK.
However, as shown in Figure \ref{map.up.clusters}, some of the most
rural areas of the UK are included in this cluster. If these areas
suffer most from first level digital divides as described in the
literature review, the low speeds in rural areas might be pulling down
the averages in other areas in this largest cluster. Also, since the
areas are clustered by their temporal profile across the working week,
the graph in Figure \ref{UpCluster} indicates that speeds in cluster
\(6\) are some of the more reliable. Table \ref{up.cluster.descr}
confirms that upload speeds in cluster \(6\) during the morning peak
from \(9:00\)-\(10:59\) were, on average, only \(4.5\)\% slower than in
the evening peak period between \(19:00\) and \(20:59\), when
entertainment purposes are likely to be using the most bandwidth.

\begin{figure}
\includegraphics[width=0.95\linewidth]{figures/map.up.clusters} \caption{\label{map.up.clusters}Upload speed clusters for LADs}\label{fig:unnamed-chunk-4}
\end{figure}

Those living in the second largest cluster -- \(9\), with \(115\) LADs
and \(21.5\) million people -- experienced aggregate mean upload speeds
of more than \(1.5\)Mb/s faster than those in cluster \(6\), a lower
ratio of standard deviation to mean, and almost the same upload speeds
in the AM Peak as the PM Peak. Indeed, the temporal profile for cluster
\(9\) in Figure \ref{UpCluster}, like that for cluster \(6\), is fairly
flat. Whilst it is likely that the large numbers of tests being
performed by the large populations in these two clusters of LADs result
in less varied averages to show in the graph, the method of shape-based
clustering suggests that these results indicate that the majority of the
population of the UK experience less temporal variation, and so belong
in these two clusters. The results also confirm that those in cluster
\(9\) experienced a consistently better service during lockdown than
those in cluster \(6\), in terms of both average speed and reliability.

Of the smaller clusters, the \(15\) LADs in clusters \(1\) and \(8\),
home to over \(1.4\) million people, not only experience fairly average
upload speeds and high ratios of standard deviation to the mean, but
also experience much lower speeds during the morning peak than the
evening peak. From the large spikes and dips shown on Figure
\ref{UpCluster}, it is likely that this poor reliability or consistency
of morning internet speeds was more noticeable to those in cluster
\(1\), whilst those in cluster \(8\) experienced the lack of reliable
service as a problem throughout the day. Meanwhile, cluster \(7\)
experiences quite similar mean speeds to cluster \(9\), and a not too
dissimilar standard deviation, but mean speeds in the morning peak are
almost \(10\)\% lower than the evening peak. Cluster \(3\) boasts speeds
almost \(1\)Mb/s higher again, but suffers from even greater slowdown in
the morning peak. In comparison, clusters \(2\) and \(5\), home to a
little less than half a million people, experience not only above
average mean speeds, but also faster speeds in the morning peak compared
to the evening. Thus, in the five LADs in these two clusters, the
temporal profile of internet use may be closer to what might have been
expected pre-pandemic.

It is worth noting here that when we apply the same methods to upload
speed data for the same time period in 2019, the spatial distribution
and composition of the nine clusters of LADs are very different -- see
Appendix \protect\hyperlink{appendix2}{2}. This difference is an
indication of the changes in the temporal profile of internet usage that
took place during the pandemic, and which would be independent of any
wider trend in improving internet services.
